### 2022-02-14 introduction/introduction.tex
more

\dropcap{H}alf sleeping, half playing Fruit Ninja on my smartphone at one of the
philosophy lectures in my high school I heard my professor exclaim with
excitement: “Mathematical truth is the highest and absolute form of truth!”.
Unbothered and uninterested, I continued playing Fruit Ninja.

Many years later I started studying various formal methods in computer science.
There I learned the value of formal, mathematical languages and finally
understood him. Not only does using a rigorous mathematical language avoid the
ambiguity of natural language, it allows us to use formal rules of reasoning to
make conclusions from initial arguments in a trustworthy manner.

@@

The initial study of this formalism and the rules to reason about statements in
it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
that there is a system of inference rules such that for any true (valid) statement a
proof of validity can be constructed in this system. Around the same time, the
first computers were leaving the imagination of the engineers and began to
automate many complicated processes.  Naturally, the question as to  whether a
computer can decide if a first-order statement is valid (i.e., a theorem) arose.
Church \cite{ac-36-fol-undecidable} and Turing \cite{tm-37-undecidable} answered
this question negatively in 1936.

>>>>

Half sleeping, half playing Fruit Ninja on my smartphone at one of the
philosophy lectures in my high school I heard my professor exclaim with
excitement: “Mathematical truth is the highest and absolute form of truth!”.
Unbothered and uninterested, I continued playing Fruit Ninja.

Many years later I started studying various formal methods in computer science.
There I learned the value of formal, mathematical language and finally
understood him. Not only does using a rigorous mathematical language avoid the
ambiguity of natural language, it allows us to use the formal rules of reasoning to
make conclusions from initial arguments in a trustworthy manner.

@@

The initial study of this formalism and the rules to reason about statements in
it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
that there is a system of inference rules (calculus) such that for any valid statement a
proof of validity can be constructed in this system. Around the same time, the
first computers were leaving the imagination of the engineers and began to
automate many complicated processes.  Naturally, the question as to  whether a
computer can decide if a first-order statement is valid (i.e., a theorem) arose.
Church \cite{ac-36-fol-undecidable} and Turing \cite{tm-37-undecidable} answered
this question negatively in 1936.

Despite this negative result, the prospect of automatically proving theorems of
first-order logic remained too enticing. In 1960 Davis and Putnam
\cite{dp-1960-dpll} described an algorithm for checking validity of a
first-order formula. As first-order logic is undecidable this algorithm
terminated only on valid formulas. Even though it was efficient enough to prove
only simple formulas, it was an impressive achievement for the time.

A bigger breakthrough happened in 1965 when Robinson introduced a calculus that
would shape the future of automatic reasoning --- the resolution calculus
\cite{ar-65-resolution}. Consisting of a single inference rule it was simple and
elegant. It was also \emph{complete}: each theorem could be proven using this
system. Furthermore, it was less explosive than Davis and Putnam's algorithm
since it introduced unification as an inference filter.

Since its introduction, resolution was improved in many ways. Strategies and
heuristics~\cite{lw-65-sos} to curb the explosion of the search space were
introduced, as well as complete, but less explosive variants of the calculus \cite{cc-73-resolution-book}. Despite this
progress, reasoning about equality of objects was still hard for these methods. This changed with introduction of
superposition in 199
